---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# qpdt

<!-- badges: start -->
<!-- badges: end -->

The goal of qpdt is to provide a package with generic pedigree tools used at Qualitas AG. 


## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("fbzwsqualitasag/qpdt")
```

## Example

This is a basic example which shows you how basic properties of a pedigree can be checked:

```{r example}
library(qpdt)
## basic example code
s_pedigree <- system.file('extdata',
                          'PopReport_SN_ohne_20210115.csv_adaptfin2.csv',
                          package = 'qprppedigree')
check_pedig_parent(ps_pedig_path = s_pedigree)
```

If you want to find whether a certain pedigree contains cycles the following statements provides an anser.

```{r cycle-check}
s_pedi_path <- system.file('extdata','data_sample2.csv', package = 'qprppedigree')
check_cycle_pedigree(ps_pedig_path = s_pedi_path)
```


```{r latest-change, echo=FALSE, results='asis'}
cat(rmdhelp::get_latest_change(), "\n")
```
