---
title: "Report Pedigree Checks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Report Pedigree Checks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(qpdt)
```


## Disclaimer
Preparatory work for creating reports based on checks of pedigrees is described.


## Goal
Provide a tool that allows for a seamless creation of a report that includes all checks and diagnosis for a given pedigree. 


## Implementation
The report is generated based on a parametrized Rmarkdown document that can be rendered into different output formats. At a later stage alternative approaches such as providing the pedigree checks in a ShinyApp might be released.


## Checks
The checks that are included into the report are determined by the available features in the R-package `qpdt`. The checks can be grouped into two classes.

1. __simple checks__ involving the verification of general properties of a pedigree
2. __extended checks__ which might require some inputs from the user to adapt the check to the specific pedigree to be checked.




